---
title: "Liver|Brain"
author: "Marta Espinosa"
date: "2024-11-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(purrr)
library(ggvenn)
library(ggpubr)
library(ggside)
library(bioseq)
# library(ggthemr)
library(ggbeeswarm)
library(ggbreak)
library(ggrepel)
library(rcartocolor)
library(UpSetR)
library(karyoploteR)
library(GenomicRanges)
library(rtracklayer)
library(Hmisc)
library(corrplot)
library(stats)
library(car)

annot = read.csv("/users/genomics/marta/TestisProject_SaraRazquin/with_TranscriptomeReconstruction/human/newReference_Resconstructed/1transcript_1gene.reconstructed.csv")
chr_annot = read.csv("/users/genomics/marta/TestisProject_SaraRazquin/with_TranscriptomeReconstruction/human/newReference_Resconstructed/gene_transcript_chr.csv")
chr_annot$gene_id = gsub("\\..*","",chr_annot$gene_id)
chr_annot$gene_id = gsub("_PAR*","",chr_annot$gene_id)
chr_annot$transcript_id = gsub("\\..*","",chr_annot$transcript_id)
annot = merge(annot, chr_annot, by=c("gene_id", "transcript_id"))
plots_wd="/projects_eg/projects/marta/TestisRestricted_Microproteins_TSA/with_TranscriptomeReconstruction/PreparationPlots/Controls"

cancers = c("BRCA","BLCA","LUAD","KIRC","PRAD","LUSC","COAD","LIHC")
tcga_projects=c("TCGA-BRCA","TCGA-LUSC","TCGA-PRAD","TCGA-KIRC","TCGA-KIRP","TCGA-LUAD","TCGA-BLCA")#,"TCGA-LIHC"]
other_projects=c("GSE102101_KIRC","GSE133624_BLCA","GSE22260_PRAD","PRJEB2449_PRAD","SRP238334_KIRC","GSE214846_LIHC","GSE229705_LUAD","TCGA_COAD","SRP107326_COAD")
manuscript_projects = c("liver_adjacent_totalRNA_LIHC","hcc_normal_totalRNA_LIHC","GSE193567_LIHC","LIHC_TCGA_LIHC")
# deleted_projects=c("GSE103001_BRCA"GSE89223_PRAD")
all_projects = c(tcga_projects,other_projects,manuscript_projects)

cancers_dir = "/users/genomics/marta/TestisProject_SaraRazquin/with_TranscriptomeReconstruction/cancers"

CTDB = read.csv("/projects_eg/projects/marta/CTdatabase_list.csv", sep="\t")
CTDB = CTDB %>% mutate(X_A = case_when(grepl("^X", chr) ~ "X",
                                       TRUE ~ "A"))
CTDB = separate_rows(CTDB, gene_name, sep = "/")
CTDB = separate_rows(CTDB, gene_name, sep = ",")

total_num_patients = data.frame(ctype = c("BRCA","BLCA","LUAD","KIRC","LUSC","PRAD","LIHC","COAD"),
                                total_n = c(109,38,179,142,49,75,182,144))

# Function to compute the required statistics
compute_stats <- function(row) {
  mean_value <- mean(row)
  median_value <- median(row)
  sd_value <- sd(row)
  q1_value <- quantile(row, 0.25)
  q3_value <- quantile(row, 0.75)
  
  return(c(mean = mean_value, median = median_value, sd = sd_value, Q1 = q1_value, Q3 = q3_value))
}

# Function to calculate log2ratio of relative frequency
calculate_log2ratio_df <- function(group1_seqs, group2_seqs) {
  all_amino_acids <- c("A","R","N","D","C","Q","E","G","H","U","L","K","M","F","P","S","T","W","Y","V", "*")
  log2ratios <- numeric(length(all_amino_acids))
  
  for (aa in all_amino_acids) {
    count_group1 <- sum(sapply(group1_seqs, function(seq) sum(unlist(strsplit(as.character(seq), "")) == aa)))
    count_group2 <- sum(sapply(group2_seqs, function(seq) sum(unlist(strsplit(as.character(seq), "")) == aa)))
    
    freq_group1 <- count_group1 / sum(sapply(group1_seqs, nchar))
    freq_group2 <- count_group2 / sum(sapply(group2_seqs, nchar))
    
    log2ratios[aa] <- log2(freq_group1 / freq_group2)
  }
  
  log2ratio_df <- data.frame(Aminoacid = all_amino_acids, log2ratio = log2ratios)
  return(log2ratio_df)
}

# "#855C75FF","#D9AF6BFF", "#AF6458FF", "#736F4CFF", "#526A83FF", "#625377FF", "#68855CFF", "#9C9C5EFF", "#A06177FF", "#8C785DFF", "#467378FF", "#7C7C7CFF"
# Extract the correlation coefficients
# res2$r
# Extract p-values
# res2$P
# A simple function to format the correlation matrix
# ++++++++++++++++++++++++++++
# flattenCorrMatrix
# ++++++++++++++++++++++++++++
# cormat : matrix of the correlation coefficients
# pmat : matrix of the correlation p-values
flattenCorrMatrix <- function(cormat, pmat) {
  ut <- upper.tri(cormat)
  data.frame(
    row = rownames(cormat)[row(cormat)[ut]],
    column = rownames(cormat)[col(cormat)[ut]],
    cor  =(cormat)[ut],
    p = pmat[ut]
  )
}
# flattenCorrMatrix(res2$r, res2$P)
```

## LiverORFs

`/home/marta/PROJECT_SCRIPTS/liverRestricted_MP/Q1.1_TestisORFs_in1.Rmd`

```{r liver, echo=F}
liver = read.csv("/projects_eg/projects/marta/TestisRestricted_Microproteins_TSA/with_TranscriptomeReconstruction/Q1_TestisORFs/human/ribORF_humanLiver_translated.csv")
print("ORFs")
table(liver %>% select(orfID, gene_id, gene_type) %>% unique() %>% pull(gene_type))
print("Genes")
table(liver %>% select(gene_id, gene_type) %>% unique() %>% pull(gene_type))
```

### Evolutionary conservation of ORFs

```{r evoLiver, echo=F}
### evolutionary conservation
conservation = read.csv("/users/genomics/marta/TestisProject_SaraRazquin/with_TranscriptomeReconstruction/EvolutionaryOrigin_MMseqs/Clustering/LIVER_allHuman_matrix_oldestAge.tsv", sep="\t")
liver = merge(liver, conservation %>% select(orfID, oldest_species), by="orfID", all.x = T)
liver[is.na(liver)] = "human"

liver = liver %>% mutate(age_categorical = case_when(oldest_species == "macaca"  ~ "primate",
                                                                 oldest_species == "human"  ~ "primate",
                                                                 oldest_species == "chicken" ~ "conserved - outgroup",
                                                                 # oldest_species == "mouse" ~ "primate",
                                                   TRUE ~ "mammal"))
liver = liver %>% mutate(coding_noncoding = case_when(gene_type == "protein_coding" ~ "coding",
                                                                TRUE ~ "noncoding"))
age_counts = liver %>% select(gene_id, coding_noncoding, age_categorical) %>% unique() %>% group_by(coding_noncoding, age_categorical) %>% count() %>% group_by(coding_noncoding) %>% mutate(proportion = n / sum(n)) 
age_counts
```

## LiverRestricted

`/home/marta/PROJECT_SCRIPTS/liverRestricted_MP/Q2_LiverRestricted.Rmd`

```{r liverRestricted, echo=F}
liverRestr = read.csv('/projects_eg/projects/marta/TestisRestricted_Microproteins_TSA/with_TranscriptomeReconstruction/liverRestricted_GTEx_translatedONLYliver.csv')
print("ORFs")
table(liverRestr %>% select(orfID, gene_id, gene_type) %>% unique() %>% pull(gene_type))
print("Genes")
table(liverRestr %>% select(gene_id, gene_type) %>% unique() %>% pull(gene_type))
```

### Evolutionary conservation of ORFs

```{r evoLiverRestr, echo=F}
## evolutionary conservation
conservation = read.csv("/users/genomics/marta/TestisProject_SaraRazquin/with_TranscriptomeReconstruction/EvolutionaryOrigin_MMseqs/Clustering/LIVER_allHuman_matrix_oldestAge.tsv", sep="\t")
liverRestr = merge(liverRestr, conservation %>% select(orfID, oldest_species), by="orfID", all.x = T)
liverRestr[is.na(liverRestr)] = "human"

liverRestr = liverRestr %>% mutate(age_categorical = case_when(oldest_species == "macaca"  ~ "primate",
                                                                 oldest_species == "human"  ~ "primate",
                                                                 oldest_species == "chicken" ~ "conserved - outgroup",
                                                                 # oldest_species == "mouse" ~ "primate",
                                                   TRUE ~ "mammal"))
liverRestr = liverRestr %>% mutate(coding_noncoding = case_when(gene_type == "protein_coding" ~ "coding",
                                                                TRUE ~ "noncoding"))
age_counts = liverRestr %>% select(gene_id, coding_noncoding, age_categorical) %>% unique() %>% group_by(coding_noncoding, age_categorical) %>% count() %>% group_by(coding_noncoding) %>% mutate(proportion = n / sum(n))
age_counts
```

## BrainORFs

`/home/marta/PROJECT_SCRIPTS/liverRestricted_MP/Q1.1_TestisORFs_in1.Rmd`

```{r brain, echo=F}
brain = read.csv("/projects_eg/projects/marta/TestisRestricted_Microproteins_TSA/with_TranscriptomeReconstruction/Q1_TestisORFs/human/ribORF_humanBrain_translated.csv")
print("ORFs")
table(brain %>% select(orfID, gene_id, gene_type) %>% unique() %>% pull(gene_type))
print("Genes")
table(brain %>% select(gene_id, gene_type) %>% unique() %>% pull(gene_type))
```

```{r evoBrain, echo=F}
## evolutionary conservation
conservation = read.csv("/users/genomics/marta/TestisProject_SaraRazquin/with_TranscriptomeReconstruction/EvolutionaryOrigin_MMseqs/Clustering/BRAIN_allHuman_matrix_oldestAge.tsv", sep="\t")
brain = merge(brain, conservation %>% select(orfID, oldest_species), by="orfID", all.x = T)
brain[is.na(brain)] = "human"

brain = brain %>% mutate(age_categorical = case_when(oldest_species == "macaca"  ~ "primate",
                                                                 oldest_species == "human"  ~ "primate",
                                                                 oldest_species == "chicken" ~ "conserved - outgroup",
                                                                 # oldest_species == "mouse" ~ "primate",
                                                   TRUE ~ "mammal"))
brain = brain %>% mutate(coding_noncoding = case_when(gene_type == "protein_coding" ~ "coding",
                                                                TRUE ~ "noncoding"))
age_counts_brain = brain %>% select(gene_id, coding_noncoding, age_categorical) %>% unique() %>% group_by(coding_noncoding, age_categorical) %>% count() %>% group_by(coding_noncoding) %>% mutate(proportion = n / sum(n))
age_counts_brain
```

## BrainRestricted

`/home/marta/PROJECT_SCRIPTS/liverRestricted_MP/Q2_BrainRestricted.Rmd`

```{r brainRestricted, echo=F}
brainRestr = read.csv('/projects_eg/projects/marta/TestisRestricted_Microproteins_TSA/with_TranscriptomeReconstruction/brainRestricted_GTEx_translatedONLYbrain.csv')
print("ORFs")
table(brainRestr %>% select(orfID, gene_id, gene_type) %>% unique() %>% pull(gene_type))
print("Genes")
table(brainRestr %>% select(gene_id, gene_type) %>% unique() %>% pull(gene_type))
```

```{r evoBrainRestr, echo=F}
## evolutionary conservation
conservation = read.csv("/users/genomics/marta/TestisProject_SaraRazquin/with_TranscriptomeReconstruction/EvolutionaryOrigin_MMseqs/Clustering/BRAIN_allHuman_matrix_oldestAge.tsv", sep="\t")
brainRestr = merge(brainRestr, conservation %>% select(orfID, oldest_species), by="orfID", all.x = T)
brainRestr[is.na(brainRestr)] = "human"

brainRestr = brainRestr %>% mutate(age_categorical = case_when(oldest_species == "macaca"  ~ "primate",
                                                                 oldest_species == "human"  ~ "primate",
                                                                 oldest_species == "chicken" ~ "conserved - outgroup",
                                                                 # oldest_species == "mouse" ~ "primate",
                                                   TRUE ~ "mammal"))
brainRestr = brainRestr %>% mutate(coding_noncoding = case_when(gene_type == "protein_coding" ~ "coding",
                                                                TRUE ~ "noncoding"))
age_counts_brainRestr = brainRestr %>% select(gene_id, coding_noncoding, age_categorical) %>% unique() %>% group_by(coding_noncoding, age_categorical) %>% count() %>% group_by(coding_noncoding) %>% mutate(proportion = n / sum(n))
age_counts_brainRestr
```

## TestisORFs

`/home/marta/PROJECT_SCRIPTS/TestisRestricted_MP/Q1.1_TestisORFs_in1.Rmd`

```{r testis, echo=F}
testis = read.csv("/projects_eg/projects/marta/TestisRestricted_Microproteins_TSA/with_TranscriptomeReconstruction/Q1_TestisORFs/human/ribORF_humanTestis_in1.csv")
print("ORFs")
table(testis %>% select(orfID, gene_id, gene_type) %>% unique() %>% pull(gene_type))
print("Genes")
table(testis %>% select(gene_id, gene_type) %>% unique() %>% pull(gene_type))
```

```{r evoTestis, echo=F}
## evolutionary conservation
conservation = read.csv("/users/genomics/marta/TestisProject_SaraRazquin/with_TranscriptomeReconstruction/EvolutionaryOrigin_MMseqs/Clustering/allHuman_matrix_oldestAge.tsv", sep="\t")
testis = merge(testis, conservation %>% select(orfID, oldest_species), by="orfID", all.x = T)
testis[is.na(testis)] = "human"

testis = testis %>% mutate(age_categorical = case_when(oldest_species == "macaca"  ~ "primate",
                                                                 oldest_species == "human"  ~ "primate",
                                                                 oldest_species == "chicken" ~ "conserved - outgroup",
                                                                 # oldest_species == "mouse" ~ "primate",
                                                   TRUE ~ "mammal"))
testis = testis %>% mutate(coding_noncoding = case_when(gene_type == "protein_coding" ~ "coding",
                                                                TRUE ~ "noncoding"))
age_counts_testis = testis %>% select(gene_id, coding_noncoding, age_categorical) %>% unique() %>% group_by(coding_noncoding, age_categorical) %>% count() %>% group_by(coding_noncoding) %>% mutate(proportion = n / sum(n))
age_counts_testis
```

## TestisRestricted

`/home/marta/PROJECT_SCRIPTS/TestisRestricted_MP/Q2_TestisRestricted_in1.Rmd`

```{r testisRestricted, echo=F}
testisRestr = read.csv('/projects_eg/projects/marta/TestisRestricted_Microproteins_TSA/with_TranscriptomeReconstruction/Q2_TestisRestricted/human/testisRestricted_GTEx_translatedONLYtestis.csv')
print("ORFs")
table(testisRestr %>% select(orfID, gene_id, gene_type) %>% unique() %>% pull(gene_type))
print("Genes")
table(testisRestr %>% select(gene_id, gene_type) %>% unique() %>% pull(gene_type))
```


```{r evotestisRestr, echo=F}
## evolutionary conservation
conservation = read.csv("/users/genomics/marta/TestisProject_SaraRazquin/with_TranscriptomeReconstruction/EvolutionaryOrigin_MMseqs/Clustering/allHuman_matrix_oldestAge.tsv", sep="\t")
testisRestr = merge(testisRestr, conservation %>% select(orfID, oldest_species), by="orfID", all.x = T)
testisRestr[is.na(testisRestr)] = "human"

testisRestr = testisRestr %>% mutate(age_categorical = case_when(oldest_species == "macaca"  ~ "primate",
                                                                 oldest_species == "human"  ~ "primate",
                                                                 oldest_species == "chicken" ~ "conserved - outgroup",
                                                                 # oldest_species == "mouse" ~ "primate",
                                                   TRUE ~ "mammal"))
testisRestr = testisRestr %>% mutate(coding_noncoding = case_when(gene_type == "protein_coding" ~ "coding",
                                                                TRUE ~ "noncoding"))
age_counts_testisRestr = testisRestr %>% select(gene_id, coding_noncoding, age_categorical) %>% unique() %>% group_by(coding_noncoding, age_categorical) %>% count() %>% group_by(coding_noncoding) %>% mutate(proportion = n / sum(n))
age_counts_testisRestr
```

### Some plots

```{r plots_translated, echo=F}
## translated
testis$tissue = "testis"
liver$tissue = "liver"
brain$tissue = "brain"

translated = rbind(testis %>% select(orfID, gene_name, coding_noncoding, tissue, oldest_species, age_categorical, gene_type),
                   liver %>% select(orfID, gene_name, coding_noncoding, tissue, oldest_species, age_categorical, gene_type),
                   brain %>% select(orfID, gene_name, coding_noncoding, tissue, oldest_species, age_categorical, gene_type))
translated_grouped = translated %>% group_by(gene_type, tissue) %>% count()
translated_grouped

ggvenn(list("Testis"= translated %>% subset(tissue == "testis") %>% pull(gene_name) %>% unique(),
                 "Liver"=translated %>% subset(tissue == "liver") %>% pull(gene_name) %>% unique(),
                 "Brain"=translated %>% subset(tissue == "brain") %>% pull(gene_name) %>% unique()),
       fill_alpha=0.5, stroke_color="gray", set_name_size = 4, text_size=3, fill_color = c("#7876B1","red","#91D1C2")) +
  ggtitle("All genetypes")
ggsave(file.path(plots_wd,"venn_tissues.png"))

## noncoding
ggvenn(list("Testis"= translated %>% subset(tissue == "testis" & coding_noncoding == "noncoding") %>% pull(gene_name) %>% unique(),
                 "Liver"=translated %>% subset(tissue == "liver" & coding_noncoding == "noncoding") %>% pull(gene_name) %>% unique(),
                 "Brain"=translated %>% subset(tissue == "brain" & coding_noncoding == "noncoding") %>% pull(gene_name) %>% unique()),
       fill_alpha=0.5, stroke_color="gray", set_name_size = 4, text_size=3, fill_color = c("#7876B1","red","#91D1C2")) +
  ggtitle("ncORFs")
ggsave(file.path(plots_wd,"venn_tissues_noncoding.png"))

## lncRNAs
ggvenn(list("Testis"= translated %>% subset(tissue == "testis" & gene_type == "lncRNA") %>% pull(gene_name) %>% unique(),
                 "Liver"=translated %>% subset(tissue == "liver" & gene_type == "lncRNA") %>% pull(gene_name) %>% unique(),
                 "Brain"=translated %>% subset(tissue == "brain" & gene_type == "lncRNA") %>% pull(gene_name) %>% unique()),
       fill_alpha=0.5, stroke_color="gray", set_name_size = 4, text_size=3, fill_color = c("#7876B1","red","#91D1C2")) +
  ggtitle("ncORFsn from lncRNAs")
ggsave(file.path(plots_wd,"venn_tissues_lncRNAs.png"))

## coding
ggvenn(list("Testis"= translated %>% subset(tissue == "testis" & coding_noncoding == "coding") %>% pull(gene_name) %>% unique(),
                 "Liver"=translated %>% subset(tissue == "liver" & coding_noncoding == "coding") %>% pull(gene_name) %>% unique(),
                 "Brain"=translated %>% subset(tissue == "brain" & coding_noncoding == "coding") %>% pull(gene_name) %>% unique()),
       fill_alpha=0.5, stroke_color="gray", set_name_size = 4, text_size=3, fill_color = c("#7876B1","red","#91D1C2")) +
  ggtitle("Coding Sequences")
ggsave(file.path(plots_wd,"venn_tissues_coding.png"))


################ EVOAGE ###############
ggplot(translated, aes(x=tissue, fill=age_categorical, fill=age_categorical)) +
  geom_bar(position="fill") +
  scale_fill_manual(values=c("primate" = "#804d5f", "mammal" = "#A06177FF", "conserved - outgroup" = "#cfb0bb")) +
  theme_minimal()
```

```{r plots_restricted, echo=F}
## translated
testisRestr$tissue = "testis"
liverRestr$tissue = "liver"
brainRestr$tissue = "brain"

translatedRestr = rbind(testisRestr %>% select(orfID, gene_name, coding_noncoding, tissue, oldest_species, age_categorical, gene_type),
                   liverRestr %>% select(orfID, gene_name, coding_noncoding, tissue, oldest_species, age_categorical, gene_type),
                   brainRestr %>% select(orfID, gene_name, coding_noncoding, tissue, oldest_species, age_categorical, gene_type))
translatedRestr_grouped = translatedRestr %>% group_by(gene_type, tissue) %>% count()
translatedRestr_grouped

ggplot(translatedRestr_grouped, aes(x=tissue, y=n, fill=gene_type)) +
  geom_bar(position="stack", stat="identity") +
  theme_minimal()
ggplot(translatedRestr_grouped, aes(x=tissue, y=n, fill=gene_type)) +
  geom_bar(position="fill", stat="identity") +
  theme_minimal()


################ EVOAGE ###############
ggplot(translatedRestr, aes(x=tissue, fill=age_categorical, fill=age_categorical)) +
  geom_bar(position="fill") +
  scale_fill_manual(values=c("primate" = "#804d5f", "mammal" = "#A06177FF", "conserved - outgroup" = "#cfb0bb")) +
  theme_minimal()
```
