---
title: "Preparation"
author: "Marta Espinosa"
date: "2024-10-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(purrr)
library(ggvenn)
library(ggpubr)
library(ggside)
library(bioseq)
# library(ggthemr)
library(ggbeeswarm)
library(ggbreak)
library(ggrepel)
library(rcartocolor)
library(UpSetR)

annot = read.csv("/users/genomics/marta/TestisProject_SaraRazquin/with_TranscriptomeReconstruction/human/newReference_Resconstructed/1transcript_1gene.reconstructed.csv")
# names(gene_transcript) = c("gene_id"gene_type"gene_name")
# gene_transcript$gene_id = gsub("\\..*"gene_transcript$gene_id)
plots_wd="/projects_eg/projects/marta/TestisRestricted_Microproteins_TSA/with_TranscriptomeReconstruction/PreparationPlots"

cancers = c("BRCA","BLCA","LUAD","KIRC","KIRP","PRAD","LUSC","COAD","LIHC")
tcga_projects=c("TCGA-BRCA","TCGA-LUSC","TCGA-PRAD","TCGA-KIRC","TCGA-KIRP","TCGA-LUAD","TCGA-BLCA")#,"TCGA-LIHC"]
other_projects=c("GSE102101_KIRC","GSE133624_BLCA","GSE22260_PRAD","PRJEB2449_PRAD","SRP238334_KIRC","GSE214846_LIHC","GSE229705_LUAD","TCGA_COAD","SRP107326_COAD")
manuscript_projects = c("liver_adjacent_totalRNA_LIHC","hcc_normal_totalRNA_LIHC","GSE193567_LIHC","LIHC_TCGA_LIHC")
# deleted_projects=c("GSE103001_BRCA"GSE89223_PRAD")
all_projects = c(tcga_projects,other_projects,manuscript_projects)

cancers_dir = "/users/genomics/marta/TestisProject_SaraRazquin/with_TranscriptomeReconstruction/cancers"

CTDB = read.csv("/projects_eg/projects/marta/CTdatabase_list.csv", sep="\t")
CTDB = CTDB %>% mutate(X_A = case_when(grepl("^X", chr) ~ "X",
                                       TRUE ~ "A"))

total_num_patients = data.frame(ctype = c("BRCA","BLCA","LUAD","KIRC","LUSC","PRAD","LIHC","COAD"),
                                total_n = c(109,38,179,142,49,75,182,144))

# Function to compute the required statistics
compute_stats <- function(row) {
  mean_value <- mean(row)
  median_value <- median(row)
  sd_value <- sd(row)
  q1_value <- quantile(row, 0.25)
  q3_value <- quantile(row, 0.75)
  
  return(c(mean = mean_value, median = median_value, sd = sd_value, Q1 = q1_value, Q3 = q3_value))
}

# Function to calculate log2ratio of relative frequency
calculate_log2ratio_df <- function(group1_seqs, group2_seqs) {
  all_amino_acids <- c("A","R","N","D","C","Q","E","G","H","U","L","K","M","F","P","S","T","W","Y","V", "*")
  log2ratios <- numeric(length(all_amino_acids))
  
  for (aa in all_amino_acids) {
    count_group1 <- sum(sapply(group1_seqs, function(seq) sum(unlist(strsplit(as.character(seq), "")) == aa)))
    count_group2 <- sum(sapply(group2_seqs, function(seq) sum(unlist(strsplit(as.character(seq), "")) == aa)))
    
    freq_group1 <- count_group1 / sum(sapply(group1_seqs, nchar))
    freq_group2 <- count_group2 / sum(sapply(group2_seqs, nchar))
    
    log2ratios[aa] <- log2(freq_group1 / freq_group2)
  }
  
  log2ratio_df <- data.frame(Aminoacid = all_amino_acids, log2ratio = log2ratios)
  return(log2ratio_df)
}

# "#855C75FF","#D9AF6BFF", "#AF6458FF", "#736F4CFF", "#526A83FF", "#625377FF", "#68855CFF", "#9C9C5EFF", "#A06177FF", "#8C785DFF", "#467378FF", "#7C7C7CFF"
```

## TestisORFs

`/home/marta/PROJECT_SCRIPTS/TestisRestricted_MP/Q1.1_TestisORFs_in1.Rmd`

```{r testis, echo=F}
testis = read.csv("/projects_eg/projects/marta/TestisRestricted_Microproteins_TSA/with_TranscriptomeReconstruction/Q1_TestisORFs/human/ribORF_humanTestis_in1.csv")
print("ORFs")
table(testis %>% select(orfID, gene_id, gene_type) %>% unique() %>% pull(gene_type))
print("Genes")
table(testis %>% select(gene_id, gene_type) %>% unique() %>% pull(gene_type))
```

## TestisRestricted - No Present in Proteome (/)BLASTP)

`/home/marta/PROJECT_SCRIPTS/TestisRestricted_MP/Q2_TestisRestricted_in1.Rmd`
`/home/marta/PROJECT_SCRIPTS/TestisRestricted_MP/Q2.1_TestisRestricted_BLASTP_ncORFSvsCDS.ipynb`

```{r testisRestricted_NoProteome, echo=F}
testisRestr = read.csv('/projects_eg/projects/marta/TestisRestricted_Microproteins_TSA/with_TranscriptomeReconstruction/Q2_TestisRestricted/human/testisRestricted_GTEx_translatedONLYtestis.csv')
print("ORFs")
table(testisRestr %>% select(orfID, gene_id, gene_type) %>% unique() %>% pull(gene_type))
print("Genes")
table(testisRestr %>% select(gene_id, gene_type) %>% unique() %>% pull(gene_type))

testisRestr = read.csv('/projects_eg/projects/marta/TestisRestricted_Microproteins_TSA/with_TranscriptomeReconstruction/Q2_TestisRestricted/human/testisRestricted_GTEx_translatedONLYtestis.noProteome.csv')
print("ORFs")
table(testisRestr %>% select(orfID, gene_id, gene_type) %>% unique() %>% pull(gene_type))
print("Genes")
table(testisRestr %>% select(gene_id, gene_type) %>% unique() %>% pull(gene_type))
```

## TumorReactivated

- Expressed 3 times more in tumor than in normal in at least 5% of the patients of at least 1 cancer type of the studied ones.

`/home/marta/PROJECT_SCRIPTS/TestisRestricted_MP/Q3.2_TestisRestricted_Tumors_in1.Rmd`
`/home/marta/PROJECT_SCRIPTS/TestisRestricted_MP/Q4_TestisRestricted_OverexpressedTumors_in1.Rmd` - TOV3x
`/home/marta/PROJECT_SCRIPTS/TestisRestricted_MP/Q4.2_Matrix.Rmd`
`/home/marta/PROJECT_SCRIPTS/TestisRestricted_MP/Q4.3_Heatmap_TTS_ctype.Rmd` - log2ratio

```{r tumorReactivated, echo=FALSE}
# tumorReact = read.csv("/projects_eg/projects/marta/TestisRestricted_Microproteins_TSA/with_TranscriptomeReconstruction/Q4_TestisRestricted_TumorSpecific/human/TOv3x_5percent_TestisRestrictedGTEx_Translated_Ctypes.csv")
tumorReact = read.csv("/projects_eg/projects/marta/TestisRestricted_Microproteins_TSA/with_TranscriptomeReconstruction/Q4_TestisRestricted_TumorSpecific/human/TOv3x_5percent_TestisRestrictedGTEx_Translated_Ctypes_log2ratio3x.csv")
# filtered_data_TOv3x_long  = read.csv("/projects_eg/projects/marta/TestisRestricted_Microproteins_TSA/with_TranscriptomeReconstruction/Q4_TestisRestricted_TumorSpecific/human/TOv3xdata_long_5percent.csv")
filtered_data_TOv3x_long  = read.csv("/projects_eg/projects/marta/TestisRestricted_Microproteins_TSA/with_TranscriptomeReconstruction/Q4_TestisRestricted_TumorSpecific/human/TOv3xlog2ratiodata_long_5percent.csv")


tumorReact = merge(tumorReact, total_num_patients, by="ctype")
print("ORFs")
table(tumorReact %>% select(orfID, gene_id, gene_type) %>% unique() %>% pull(gene_type))
print("Genes")
table(tumorReact %>% select(gene_id, gene_type) %>% unique() %>% pull(gene_type))

print("Per cancertype")
ctypes_tumorReact = tumorReact %>% mutate(coding_ncoding = case_when(gene_type == "protein_coding" ~ "protein_coding",
                                                                     TRUE ~ "noncoding")) %>% select(gene_id, ctype, coding_ncoding) %>% unique()
table(ctypes_tumorReact$coding_ncoding, ctypes_tumorReact$ctype)
table(ctypes_tumorReact$ctype)
```

```{r correlation_npatients_nTSTR, echo=F}
tumorReact %>% select(gene_id, total_n, ctype) %>% group_by(ctype, total_n) %>% count() %>%
  ggplot(., aes(x=total_n, y=n, color=ctype)) +
  geom_point(stat="identity", size=5) +
  geom_text(aes(label=ctype), hjust = -0.3, size=3, fontface = "bold") +
  scale_x_continuous(limits=c(25,200)) +
  labs(x="Total number of patients",
       y="Number of TS/TR genes",
       title="Correlation between TS/TR and\nthe number of patients",
       color="CancerType") +
    scale_color_manual(values=c("BRCA"="#E69F00", "BLCA"="#56A3A6", "LUAD"="#0079E3", "KIRC"="#F0C27B", "LUSC"="#0072B2", "PRAD"="#D55E00", "LIHC"="#CC79A7", "COAD"="#8B7355")) +
  theme_minimal() +
  theme(legend.position="none")

ggsave(file.path(plots_wd, "PNG/TSTR_numpatients_scatter.png"), height=5.08, width=4.80)

```

## Correlation testis expression vs tumor expression?

```{r testis_tumor_TPMs, echo=F}
# raw_counts = read.table("/users/genomics/saraa/projectTestis/featureCounts/results_geneID/featureCounts_geneID.txt header=TRUE)
testis_TPMs = read.csv("/users/genomics/marta/TestisProject_SaraRazquin/with_TranscriptomeReconstruction/human/featureCounts_gffcompare/table_of_counts_TPMs_withLength.csv", header=TRUE)
names(testis_TPMs)[1] = "transcript_id"
## For novels, remove those shorter than 300 pb
testis_TPMs = testis_TPMs %>%
  filter(!(startsWith(transcript_id, "TCONS") & Length < 300))

testis_TPMs = merge(testis_TPMs, annot, by="transcript_id", all.x=T)
testis_TPMs = testis_TPMs %>% unique()
testis_TPMs = testis_TPMs[,!grepl("brain", names(testis_TPMs))]
testis_TPMs = testis_TPMs[,!grepl("liver", names(testis_TPMs))]
testis_TPMs = testis_TPMs %>% select(-Length)

## Select only numeric columns
numeric_df = testis_TPMs[, sapply(testis_TPMs, is.numeric)]
## Apply the function to each row
stats_per_row = t(apply(numeric_df, 1, compute_stats))
stats_df = as.data.frame(stats_per_row)

testis_TPMs_stats = cbind(testis_TPMs %>% select(transcript_id, gene_id, gene_type, gene_name), stats_df)
testis_TPMs_stats %>% head
names(testis_TPMs_stats) = c("transcript_id","gene_id","gene_type","gene_name","mean_testis","median_testis","sd_testis","Q1.25%_testis","Q3.75%_testis")


### Testis-Tumor
testis_tumor_TPMs = merge(filtered_data_TOv3x_long, testis_TPMs_stats, by=c("transcript_id","gene_id","gene_type","gene_name"))
## remove the zeros, show mean of patients for which it is Overexpressed
testis_tumor_TPMs = testis_tumor_TPMs %>% subset(normal_tumor == "tumor") %>% subset(forced_TPM > 0) 
testis_tumor_TPMs = testis_tumor_TPMs %>% group_by(gene_name, gene_type, ctype, mean_testis, median_testis) %>% mutate(mean_tumor = mean(forced_TPM),
                                                                                          median_tumor = median(forced_TPM))
testis_tumor_TPMs = testis_tumor_TPMs %>% select(gene_name, gene_type, mean_tumor, median_tumor, mean_testis, median_testis) %>% unique()

### MEAN tumor zeros removed
testis_tumor_TPMs %>% mutate(coding_noncoding = case_when(gene_type == "protein_coding" ~ "protein_coding",
                                                          TRUE ~ "noncoding")) %>%
  ggplot(., aes(x=mean_tumor, y=mean_testis, color=coding_noncoding)) +
  geom_point() +
  geom_text_repel(aes(label=gene_name)) +
  labs(x="Mean TPM Tumor Samples (zeros removed)",
       y="Mean TPM Testis Samples",
       title="Correlation expression Testis vs Tumor") +
  scale_x_continuous(limits=c(0,275)) +
  scale_color_manual(values=c("#DFA398","#3B3960")) +
  theme_minimal() +
  facet_wrap(~ ctype)

### MEAN tumor zeros removed
testis_tumor_TPMs %>% mutate(coding_noncoding = case_when(gene_type == "protein_coding" ~ "protein_coding",
                                                          TRUE ~ "noncoding")) %>%
  ggplot(., aes(x=log(mean_tumor), y=log(mean_testis), color=coding_noncoding)) +
  geom_point() +
  # geom_smooth(method = "lm", se = FALSE, aes(color=coding_noncoding)) +
   # stat_cor(label.y = c(5.2, 4.8, 4.3, 3.8), label.x = c(1,1,1,1), aes(group=coding_noncoding, color = coding_noncoding), size=3) +
  labs(x="log(Mean TPM) Tumor Samples (zeros removed)",
       y="log(Mean TPM) Testis Samples",
       title="Correlation expression Testis vs Tumor") +
  scale_y_continuous(limits=c(-2.5,6)) +
  scale_x_continuous(limits=c(-2.5,6)) +
  geom_vline(xintercept=0, size=.2) +
  geom_hline(yintercept=0, size=.2) +
  scale_color_manual(values=c("#DFA398","#3B3960")) +
  theme_minimal() +
  theme(legend.position="top") +
  facet_wrap(~ ctype)
ggsave(file.path(plots_wd, "PNG/TSTR_TestisExpr_TumorExpr.png"), width=7, height=7)
```

## Comparisons, from TestisSpecific, which are Reactivated and which are not

### Length

```{r comparisons, echo=F}
testisRestr = testisRestr %>% 
  mutate(tumorReactivated = case_when(gene_id %in% tumorReact$gene_id ~ "Reactivated",
                                      TRUE ~ "NoReactivated")) %>%
  mutate(coding_noncoding = case_when(gene_type == "protein_coding" ~ "protein_coding",
                                                          TRUE ~ "noncoding"))
```

```{r length, echo=F}
### length
## KruskalWallis
testisRestr %>% select(gene_id, coding_noncoding, length_aa, tumorReactivated) %>% unique() %>%
  ggplot(., aes(y=length_aa, x=coding_noncoding, fill=tumorReactivated)) +
  geom_boxplot() +
  stat_compare_means(label="p.signif") +
  labs(x="",
       y="Length (AA)",
       title = "Length differences between TS/TR and TS/nonTR",
       fill="Reactivation\nin tumor") +
  scale_fill_manual(values=c("#D9AF6BFF","#625377FF")) +
  scale_y_continuous(trans="log10") +
  theme_minimal()
ggsave(file.path(plots_wd, "PNG/TSTR_Length_boxplot.png"), width=6.28, height=5.82)
```

## Translation efficiency

```{r TE, echo=F}
### translation efficiency
testis_TE = read.csv("/projects_eg/projects/marta/TestisRestricted_Microproteins_TSA/with_TranscriptomeReconstruction/Q1_TestisORFs/human/mean_translationEfficiency_testis.csv")
testisRestr = merge(testisRestr, testis_TE %>% select(orfID, mean_TE), by="orfID")

testisRestr %>% select(gene_id, coding_noncoding, mean_TE, tumorReactivated) %>% unique() %>%
  ggplot(., aes(y=mean_TE, x=coding_noncoding, fill=tumorReactivated)) +
  geom_boxplot() +
  stat_compare_means(label="p.signif") +
  scale_y_continuous(trans="log10") +
  labs(x="",
       y="Mean Translation Efficiency",
       title = "TE differences between TS/TR and TS/nonTR",
       fill="Reactivation\nin tumor") +
  scale_fill_manual(values=c("#D9AF6BFF","#625377FF")) +
  theme_minimal()
ggsave(file.path(plots_wd, "PNG/TSTR_TE_boxplot.png"), width=6.28, height=5.82)

```

## Evolutionary conservation of ORFs

```{r evo, echo=F}
### evolutionary conservation
conservation = read.csv("/users/genomics/marta/TestisProject_SaraRazquin/with_TranscriptomeReconstruction/EvolutionaryOrigin_MMseqs/Clustering/allHuman_matrix_oldestAge.tsv", sep="\t")
testisRestr = merge(testisRestr, conservation %>% select(orfID, oldest_species), by="orfID", all.x = T)
testisRestr[is.na(testisRestr)] = "human"

testisRestr = testisRestr %>% mutate(age_categorical = case_when(oldest_species == "macaca"  ~ "primate",
                                                                 oldest_species == "human"  ~ "primate",
                                                                 # oldest_species == "chicken" ~ "conserved - outgroup",
                                                                 # oldest_species == "mouse" ~ "primate",
                                                   TRUE ~ "mammal"))

age_counts = testisRestr %>% select(gene_id, coding_noncoding, age_categorical, tumorReactivated) %>% unique() %>% group_by(coding_noncoding, tumorReactivated, age_categorical) %>% count() %>% group_by(coding_noncoding, tumorReactivated) %>% mutate(proportion = n / sum(n)) 
age_counts = as.data.frame(age_counts)
ggplot(age_counts, aes(y=proportion, x=factor(age_categorical, levels=c("primate","mammal","conserved - outgroup")), fill=tumorReactivated, group=tumorReactivated)) +
  geom_bar(stat="identity", position="dodge") +
  labs(x="Conservation level",
       y="Proportion",
       title="Evolutionary Conservation of TS/TR vs TS/nonTR") +
  scale_fill_manual(values=c("#D9AF6BFF","#625377FF")) +
  theme_minimal() +
  facet_grid( ~ coding_noncoding)
ggsave(file.path(plots_wd, "PNG/TSTR_Conservation_bar.png"), width=8.11, height=4.82)

# Subset the data for protein coding genes
protein_coding_data <- subset(age_counts, coding_noncoding == "protein_coding")

# Create a 2x2 matrix for 'primate' vs 'mammal' in Reactivated vs NoReactivated
contingency_table <- matrix(c(
  sum(protein_coding_data$n[protein_coding_data$tumorReactivated == "Reactivated" & protein_coding_data$age_categorical == "primate"]),
  sum(protein_coding_data$n[protein_coding_data$tumorReactivated == "NoReactivated" & protein_coding_data$age_categorical == "primate"]),
  sum(protein_coding_data$n[protein_coding_data$tumorReactivated == "Reactivated" & protein_coding_data$age_categorical == "mammal"]),
  sum(protein_coding_data$n[protein_coding_data$tumorReactivated == "NoReactivated" & protein_coding_data$age_categorical == "mammal"])
), nrow = 2, byrow = TRUE)

# Add proper labels to the matrix for clarity
dimnames(contingency_table) <- list(
  "Reactivated" = c("Reactivated", "NoReactivated"),
  "Age Category" = c("Primate", "Mammal")
)

# Perform Fisher's exact test
fisher.test(contingency_table)

testisRestr = testisRestr %>% mutate(age_categorical = case_when(oldest_species == "macaca"  ~ "primate",
                                                                 oldest_species == "human"  ~ "primate",
                                                                 oldest_species == "chicken" ~ "conserved - outgroup",
                                                                 # oldest_species == "mouse" ~ "primate",
                                                   TRUE ~ "mammal"))
age_counts = testisRestr %>% select(gene_id, coding_noncoding, age_categorical, tumorReactivated) %>% unique() %>% group_by(coding_noncoding, tumorReactivated, age_categorical) %>% count() %>% group_by(coding_noncoding, tumorReactivated) %>% mutate(proportion = n / sum(n)) 
age_counts = as.data.frame(age_counts)

ggplot(age_counts %>% subset(coding_noncoding == "protein_coding"), aes(x = tumorReactivated, y = proportion, fill = age_categorical)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values=c("primate" = "#A06177FF", "mammal" = "#9C9C5EFF", "conserved - outgroup" = "#467378FF")) +
  labs(x="Reactivation in tumor",
       y="Proportion",
       title="Proportion of ORF Conservation\ndepending on Tumor Reactivation") +
  # geom_text(aes(label=round(proportion,2)), position = "stack") +
  theme_minimal() 
ggsave(file.path(plots_wd, "PNG/TSTR_Conservation_bar_coding.png"), width=5.78, height=5.82)

ggplot(age_counts, aes(x = tumorReactivated, y = proportion, fill = age_categorical)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values=c("primate" = "#A06177FF", "mammal" = "#9C9C5EFF", "conserved - outgroup" = "#467378FF")) +
  labs(x="Reactivation in tumor",
       y="Proportion",
       title="Proportion of ORF Conservation\ndepending on Tumor Reactivation") +
  theme(legend.position = "top") +
  # geom_text(aes(label=round(proportion,2)), position = "stack") +
  theme_minimal() +
  facet_wrap(~ coding_noncoding)
ggsave(file.path(plots_wd, "PNG/TSTR_Conservation_bar.png"), width=7.78, height=5.82)

ggplot(testisRestr, aes(x=age_categorical, y=length_aa, color=age_categorical)) +
  geom_quasirandom(size=1) +
  scale_color_manual(values=c("primate" = "#A06177FF", "mammal" = "#9C9C5EFF", "conserved - outgroup" = "#467378FF")) +
  scale_y_continuous(trans="log10") +
  labs(x="Evolutionary Conservation of ORFs",
       y="Length (AA)",
       title="Relationship between Conservation and ORF length") +
  theme_minimal() +
  theme(legend.position = "top") +
  facet_wrap(~ coding_noncoding) +
  stat_compare_means(comparisons = list(c("mammal", "primate"),c("mammal", "conserved - outgroup"),c("conserved - outgroup", "primate")), label="p.signif", method = "wilcox.test", paired=F)
ggsave(file.path(plots_wd, "PNG/Conservation_Length.png"), width=6.21, height=5.25)

ggplot(testisRestr, aes(x=age_categorical, y=length_aa, color=age_categorical)) +
  geom_quasirandom(size=1) +
  scale_color_manual(values=c("primate" = "#A06177FF", "mammal" = "#9C9C5EFF", "conserved - outgroup" = "#467378FF")) +
  scale_y_continuous(trans="log10") +
  labs(x="Evolutionary Conservation of ORFs",
       y="Length (AA)",
       title="Relationship between Conservation and ORF length") +
  theme_minimal() +
  theme(legend.position = "top") +
  facet_grid(tumorReactivated ~ coding_noncoding) +
  stat_compare_means(comparisons = list(c("mammal", "primate"),c("mammal", "conserved - outgroup"),c("conserved - outgroup", "primate")), label="p.signif", method = "wilcox.test", paired=F)
ggsave(file.path(plots_wd, "PNG/TSTR_Conservation_Length.png"), width=6.21, height=6.25)
```

## Aminoacid Composition

First aa removed

```{r composition, echo=F}
### aminoacid composition
hph=c("A","V","L","I","M","F","Y","W")
polar=c("S","T","N","Q")
positively_charged=c("R","H","K")
negatively_charged=c("D","E")
special_cases=c("C","G","P")

aacomposition_testisRestr = testisRestr %>% select(tumorReactivated, orfID, coding_noncoding, ORFpep) %>% unique()
## remove starting codon
aacomposition_testisRestr$ORFpep = substring(aacomposition_testisRestr$ORFpep, 2)
aacomposition_testisRestr$ORFpep = gsub("\\*","",as.character(aacomposition_testisRestr$ORFpep))


#### non-coding
nc_React = aacomposition_testisRestr %>% subset(coding_noncoding == "noncoding") %>% subset(tumorReactivated == "Reactivated")
nc_React = as.character(nc_React$ORFpep)
nc_NoReact = aacomposition_testisRestr %>% subset(coding_noncoding == "noncoding") %>% subset(tumorReactivated != "Reactivated")
nc_NoReact = as.character(nc_NoReact$ORFpep)

log2ratios_noncoding = calculate_log2ratio_df(nc_React, nc_NoReact)

log2ratios_noncoding %>% 
  mutate(aa_type = case_when(Aminoacid %in% hph ~ "hidrophobic",
                             Aminoacid %in% polar ~ "polar",
                             Aminoacid %in% positively_charged ~ "positive",
                             Aminoacid %in% negatively_charged ~ "negative",
                             Aminoacid %in% special_cases ~ "special")) %>%
  subset(Aminoacid != "*") %>% 
  drop_na() %>%
ggplot(., aes(x=reorder(Aminoacid,log2ratio), y=log2ratio, fill = aa_type)) +
  geom_bar(stat = "identity") +
  labs(y="Log2Ratio\n(Reactivated/No Reactivated)",
       x="Aminoacid",
       title="Relative enrichement of aminoacids in\nNon-coding genes Reactivated in cancer") +
  scale_fill_manual(values = c("#462255FF", "#FF8811FF", "#9DD9D2FF","#046E8FFF", "#D44D5CFF")) +
  theme_classic()
ggsave(file.path(plots_wd, "PNG/TSTR_AAComposition_noncoding.png"), width=8.38, height=3.82)

#### coding
c_React = aacomposition_testisRestr %>% subset(coding_noncoding == "protein_coding") %>% subset(tumorReactivated == "Reactivated")
c_React = as.character(c_React$ORFpep)
c_NoReact = aacomposition_testisRestr %>% subset(coding_noncoding == "protein_coding") %>% subset(tumorReactivated != "Reactivated")
c_NoReact = as.character(c_NoReact$ORFpep)

log2ratios_coding = calculate_log2ratio_df(c_React, c_NoReact)
log2ratios_coding %>% 
  mutate(aa_type = case_when(Aminoacid %in% hph ~ "hidrophobic",
                             Aminoacid %in% polar ~ "polar",
                             Aminoacid %in% positively_charged ~ "positive",
                             Aminoacid %in% negatively_charged ~ "negative",
                             Aminoacid %in% special_cases ~ "special")) %>%
  subset(Aminoacid != "*") %>% drop_na() %>% 
ggplot(., aes(x=reorder(Aminoacid,log2ratio), y=log2ratio, fill = aa_type)) +
  geom_bar(stat = "identity") +
  labs(y="Log2Ratio\n(Reactivated/No Reactivated)",
       x="Aminoacid",
       title="Relative enrichement of aminoacids in\nProteinCoding genes Reactivated in cancer") +
  scale_fill_manual(values = c("#462255FF", "#FF8811FF", "#9DD9D2FF","#046E8FFF", "#D44D5CFF")) +
  theme_classic()
ggsave(file.path(plots_wd, "PNG/TSTR_AAComposition_proteincoding.png"), width=8.38, height=3.82)

### coding vs noncoding
coding_AA = aacomposition_testisRestr %>% subset(coding_noncoding == "protein_coding")
coding_AA = as.character(coding_AA$ORFpep)
noncoding_AA = aacomposition_testisRestr %>% subset(coding_noncoding == "noncoding")
noncoding_AA = as.character(noncoding_AA$ORFpep)

log2ratios_codingnoncoding = calculate_log2ratio_df(noncoding_AA, coding_AA)
log2ratios_codingnoncoding %>% 
  mutate(aa_type = case_when(Aminoacid %in% hph ~ "hidrophobic",
                             Aminoacid %in% polar ~ "polar",
                             Aminoacid %in% positively_charged ~ "positive",
                             Aminoacid %in% negatively_charged ~ "negative",
                             Aminoacid %in% special_cases ~ "special")) %>%
  subset(Aminoacid != "*") %>% drop_na() %>% 
ggplot(., aes(x=reorder(Aminoacid,log2ratio), y=log2ratio, fill = aa_type)) +
  geom_bar(stat = "identity") +
  labs(y="Log2Ratio\n(Non-Coding/Coding)",
       x="Aminoacid",
       title="Relative enrichement of aminoacids in ncORFs wrt CDS") +
  scale_fill_manual(values = c("#462255FF", "#FF8811FF", "#9DD9D2FF","#046E8FFF", "#D44D5CFF")) +
  theme_classic()
ggsave(file.path(plots_wd, "PNG/AAComposition_coding_vs_noncoding.png"), width=8.38, height=3.82)

```

## Tumor-specific antigens contribution

- Cryptic DB

- Beijer et al.

- Erhard et al.

- IEAtlas

- Chong et al.

- Ouspenskaia et al.

```{r neoantigens, echo=F}
## Reactivated in tumor
# tumorReact_TSA = read.csv("/projects_eg/projects/marta/TestisRestricted_Microproteins_TSA/with_TranscriptomeReconstruction/Q5_immunopeptidomics/human/TOv3x_5percent_TestisRestrictedGTEx_Translated_Ctypes_immuno.csv")
tumorReact_TSA = read.csv("/projects_eg/projects/marta/TestisRestricted_Microproteins_TSA/with_TranscriptomeReconstruction/Q5_immunopeptidomics/human/TOv3x_5percent_TestisRestrictedGTEx_Translated_Ctypes_log2ratio3x_immuno.csv")

print("How many ORFs contain at least 1 TSA evidenced by the literature? | Tumor Reactivated")
print(length(unique(tumorReact_TSA$matching_ORFpep)))
print("From how many total ORFs?")
print(length(unique(tumorReact$ORFpep)))
print(paste0("Proportion = ",length(unique(tumorReact_TSA$matching_ORFpep))/length(unique(tumorReact$ORFpep))*100,"%"))

## TestisExpressed in tumor
testisRestr_TSA = read.csv("/projects_eg/projects/marta/TestisRestricted_Microproteins_TSA/with_TranscriptomeReconstruction/Q5_immunopeptidomics/human/TestisRestrictedGTEx_Translated_ONLYintestis_immuno.csv")
print("How many ORFs contain at least 1 TSA evidenced by the literature? | TestisRestricted")
print(length(unique(testisRestr_TSA$matching_ORFpep)))
print("From how many total ORFs?")
print(length(unique(testisRestr$ORFpep)))
print(paste0("Proportion = ",length(unique(testisRestr_TSA$matching_ORFpep))/length(unique(testisRestr$ORFpep))*100,"%"))

## how many of each condition
numbers_TSA = data.frame("TumorReactivated" = c("Reactivated","NoReactivated"),
                         "ORFs_with_evidence" = c(length(unique(tumorReact_TSA$matching_ORFpep)), length(unique(testisRestr_TSA$matching_ORFpep))),
                         "totalORFs" = c(length(unique(tumorReact$ORFpep)), length(unique(testisRestr$ORFpep))))
numbers_TSA = numbers_TSA %>%
  mutate(proportion = numbers_TSA$ORFs_with_evidence / numbers_TSA$totalORFs * 100) %>% ## compute proportion of evidenced
  mutate(ORFs_without_evidence = numbers_TSA$totalORFs - numbers_TSA$ORFs_with_evidence)
  
numbers_TSA %>%
  select(TumorReactivated, ORFs_with_evidence, ORFs_without_evidence) %>%
  pivot_longer(cols = -TumorReactivated, names_to = "Condition", values_to = "ORFs") %>%
  group_by(TumorReactivated) %>%
  mutate(Percentage = ORFs / sum(ORFs) * 100,
         Cumulative_ORFs = cumsum(ORFs),
         LabelPosition = Cumulative_ORFs - (ORFs / 2)) %>% # Position in the middle of each slice based on ORFs
  ggplot() + 
  geom_bar(aes(x = 1, y = Percentage, fill = Condition), stat = "identity") +
  coord_polar("y", start = 0) +
  ggtitle("ORFs with immunopeptidomics evidence") +
  theme_void() +  
    scale_fill_manual(values=c("#855C75FF","#D9AF6BFF")) +
  theme(legend.position = "top") +
  facet_wrap(~ TumorReactivated, scales = "free") 
ggsave(file.path(plots_wd, "PNG/TSTR_ORFs_with_Immuno.png"))

testisRestr = merge(testisRestr, testisRestr_TSA %>% select(gene_name, Peptide, ORFpep, source), by=c("gene_name","ORFpep"), all.x=T)
testisRestr[is.na(testisRestr)] = "None"

## how many peptides per ORF?
pep_x_orf = testisRestr %>% select(orfID, Peptide, gene_name, coding_noncoding, tumorReactivated, length_aa) %>% unique() %>% group_by(orfID, coding_noncoding, gene_name, tumorReactivated, length_aa) %>% count()

## all non-coding have only 1 peptide per ORF
ggplot(pep_x_orf %>% subset(coding_noncoding == "protein_coding"), aes(x=n, y=length_aa)) +
  geom_point() +
  geom_text_repel(aes(label=gene_name), size=3, color="black") +
  scale_y_continuous(trans="log10") +
  labs(x="Number of evidenced peptides",
       y="Length (AA)",
       title="Number of immunopeptides per ORF vs length ORFs") +
  theme_minimal() 
ggsave(file.path(plots_wd, "PNG/TSTR_ORFs_vs_Length.png"), width=6.91, height=5.90)

## at peptide level
TSA = testisRestr %>% select(orfID, Peptide, source, gene_name, coding_noncoding, tumorReactivated) %>% unique()
TSA$value = 1
TSA = TSA %>% pivot_wider(names_from = "source", values_from = "value")
TSA[is.na(TSA)] = 0
TSA$sum = rowSums(TSA[,c(6:9)])

## can it be that an ORF is validated by two sources, even though it is a different peptide?
TSA_ORF = testisRestr %>% select(orfID, source, gene_name, coding_noncoding, tumorReactivated) %>% unique()
TSA_ORF$value = 1
TSA_ORF = TSA_ORF %>% pivot_wider(names_from = "source", values_from = "value")
TSA_ORF[is.na(TSA_ORF)] = 0
TSA_ORF$sum = rowSums(TSA_ORF[,c(5:8)])

sources = testisRestr %>% select(gene_name, orfID, tumorReactivated, source) %>% unique()
table(sources$tumorReactivated, sources$source)
## which are the noReactivated with immunopeptidomics evidence?

NoReactivatedTSAexamples = testisRestr %>% subset(tumorReactivated == "NoReactivated" & source != "None") %>% select(transcript_id, gene_name, orfID, length_aa, source, gene_type, age_categorical) %>% unique()

testisRestr %>% 
  select(gene_name, orfID, tumorReactivated, source, coding_noncoding) %>% 
  unique() %>% 
  group_by(tumorReactivated, source, coding_noncoding) %>% count() %>%
  mutate(Immuno = case_when(source == "None" ~ "None",
                            TRUE ~ "Immunovalidated")) %>%
  ggplot(., aes(x=tumorReactivated, y=n, fill=source, group = Immuno)) +
  geom_bar(stat="identity", position = "fill") +
  scale_fill_manual(values=c("deBeijer" = "#8C785DFF",
                             "IEAtlas" = "#AF6458FF",
                             "nuORFdb" = "#625377FF",
                             "DiazCamarena" = "#9C9C5EFF",
                             "None" = "#DCDCDC")) +
  labs(x="Tumor Reactivation",
       fill="Source of\nimmunopeptidomics evidence",
       y="Proportion of Tumor-specific antigens",
       title="Number of immunovalidated TSA in TSTR") +
  theme_minimal() +
  facet_wrap(~ coding_noncoding)
ggsave(file.path(plots_wd, "PNG/TSTR_TSAburden_immuno.png"), width=7, height=6.18)

##  How are the examples in terms of tumorExpression
# unfiltered_allfc = read.csv("/projects_eg/projects/marta/TestisRestricted_Microproteins_TSA/with_TranscriptomeReconstruction/Q3_TestisRestrictedTumors/human/unfiltered_tablesofcounts_long.csv")
# NoReactivatedTSAexamples_full = unfiltered_allfc %>% subset(transcript_id %in% NoReactivatedTSAexamples$transcript_id)
# NoReactivatedTSAexamples_stats = NoReactivatedTSAexamples_full %>% group_by(transcript_id, normal_tumor) %>% 
#   mutate(mean = mean(TPM),
#          median = median(TPM),
#          max = max(TPM)) %>% select(transcript_id, normal_tumor, mean, median, max) %>% unique()

cases = testisRestr %>% 
  select(gene_name, orfID, tumorReactivated, source, coding_noncoding) %>% 
  unique() %>% subset(source != "None") %>% select(-c(source, orfID)) %>% unique()
cases

to_venn = testisRestr %>% 
  select(gene_name, orfID, tumorReactivated, source, coding_noncoding) %>% 
  unique() %>% subset(source != "None")
table(to_venn$source, to_venn$coding_noncoding)

ggvenn(list("IEAtlas" = to_venn %>% subset(coding_noncoding == "protein_coding" & source == "IEAtlas" & tumorReactivated == "Reactivated") %>% pull(orfID),
            "nuORFdb" = to_venn %>% subset(coding_noncoding == "protein_coding" & source == "nuORFdb" & tumorReactivated == "Reactivated") %>% pull(orfID),
            "DiazCamarena" = to_venn %>% subset(coding_noncoding == "protein_coding" & source == "DiazCamarena" & tumorReactivated == "Reactivated") %>% pull(orfID),
            "deBeijer" = to_venn %>% subset(coding_noncoding == "protein_coding" & source == "deBeijer" & tumorReactivated == "Reactivated") %>% pull(orfID)),
       fill_alpha=0.5, stroke_color="gray", set_name_size = 4, text_size=3, 
       fill_color = c("#AF6458FF","#625377FF","#9C9C5EFF","#8C785DFF","#DCDCDC")) +
  ggtitle("Protein-Coding ORFs Reactivated in Tumors")
ggsave(file.path(plots_wd, "PNG/venn_TSTR_CDS.png"), width=7, height=6.18)


ggvenn(list("IEAtlas" = to_venn %>% subset(coding_noncoding == "protein_coding" & source == "IEAtlas" & tumorReactivated == "NoReactivated") %>% pull(orfID),
            "nuORFdb" = to_venn %>% subset(coding_noncoding == "protein_coding" & source == "nuORFdb" & tumorReactivated == "NoReactivated") %>% pull(orfID),
            "DiazCamarena" = to_venn %>% subset(coding_noncoding == "protein_coding" & source == "DiazCamarena" & tumorReactivated == "NoReactivated") %>% pull(orfID),
            "deBeijer" = to_venn %>% subset(coding_noncoding == "protein_coding" & source == "deBeijer" & tumorReactivated == "NoReactivated") %>% pull(orfID)),
       fill_alpha=0.5, stroke_color="gray", set_name_size = 4, text_size=3, 
       fill_color = c("#AF6458FF","#625377FF","#9C9C5EFF","#8C785DFF","#DCDCDC")) +
  ggtitle("Protein-Coding ORFs NoReactivated in Tumors")
ggsave(file.path(plots_wd, "PNG/venn_TSnoTR_CDS.png"), width=7, height=6.18)


ggvenn(list("IEAtlas" = to_venn %>% subset(coding_noncoding == "noncoding" & source == "IEAtlas" & tumorReactivated == "Reactivated") %>% pull(orfID),
            "nuORFdb" = to_venn %>% subset(coding_noncoding == "noncoding" & source == "nuORFdb" & tumorReactivated == "Reactivated") %>% pull(orfID),
            "DiazCamarena" = to_venn %>% subset(coding_noncoding == "noncoding" & source == "DiazCamarena" & tumorReactivated == "Reactivated") %>% pull(orfID),
            "deBeijer" = to_venn %>% subset(coding_noncoding == "noncoding" & source == "deBeijer" & tumorReactivated == "Reactivated") %>% pull(orfID)),
       fill_alpha=0.5, stroke_color="gray", set_name_size = 4, text_size=3, 
       fill_color = c("#AF6458FF","#625377FF","#9C9C5EFF","#8C785DFF","#DCDCDC")) +
  ggtitle("Non-Canonical ORFs Reactivated in Tumors")
ggsave(file.path(plots_wd, "PNG/venn_TSTR_ncORFs.png"), width=7, height=6.18)
```

## Selection of TS/TR

```{r TSTR, echo=F}
TSTR = testisRestr %>% subset(tumorReactivated == "Reactivated") %>% select(-X)
TSTR_ORFs = merge(TSTR, tumorReact %>% select(orfID, total_n, percentage_num_patients_overexpr, num_patients_overexpr, ctype), by=c("orfID"), all.x=T)
TSTR_genes = merge(TSTR %>% select(-c(orfID, ORFpep, length_aa, length, start_codon, mean_TE, TranslatedLiver, TranslatedBrain, orfType)) %>% unique(), tumorReact %>% select(gene_id, total_n, percentage_num_patients_overexpr, num_patients_overexpr, ctype), by=c("gene_id"), all.x=T)

TSTR_genes_fewinfo = TSTR_genes %>% select(gene_id, gene_name, coding_noncoding, total_n, percentage_num_patients_overexpr, num_patients_overexpr, ctype) %>% unique()

## ProteinCoding
to_upset_Coding = TSTR_genes_fewinfo %>% subset(coding_noncoding == "protein_coding") %>% select(ctype, gene_name) %>% unique()
to_upset_Coding$presence = 1
to_upset_Coding = to_upset_Coding %>% pivot_wider(names_from = "ctype", values_from = "presence")
to_upset_Coding[is.na(to_upset_Coding)] = 0
to_upset_Coding = as.data.frame(to_upset_Coding)
rownames(to_upset_Coding) = to_upset_Coding$gene_name
to_upset_Coding$gene_name = NULL

upset(to_upset_Coding,
      sets=colnames(to_upset_Coding), main.bar.color = "#CC79A7"
)
to_upset_Coding_sums = to_upset_Coding
to_upset_Coding_sums$sum = rowSums(to_upset_Coding_sums)


## Non-Coding
to_upset_nonCoding = TSTR_genes_fewinfo %>% subset(coding_noncoding == "noncoding") %>% select(ctype, gene_name) %>% unique()
to_upset_nonCoding$presence = 1
to_upset_nonCoding = to_upset_nonCoding %>% pivot_wider(names_from = "ctype", values_from = "presence")
to_upset_nonCoding[is.na(to_upset_nonCoding)] = 0
to_upset_nonCoding = as.data.frame(to_upset_nonCoding)
rownames(to_upset_nonCoding) = to_upset_nonCoding$gene_name
to_upset_nonCoding$gene_name = NULL

upset(to_upset_nonCoding,
      sets=colnames(to_upset_nonCoding), main.bar.color = "#0090B2"
)

to_upset_nonCoding_sums = to_upset_nonCoding
to_upset_nonCoding_sums$sum = rowSums(to_upset_nonCoding_sums)
```
