---
title: "MMSeqs2 Conservation"
author: "Marta Espinosa"
date: "2024-10-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(purrr)
library(ggvenn)
library(ggpubr)
library(ggside)
library(bioseq)
library(ggbeeswarm)
library(ggbreak)
library(pheatmap)
library(ggbreak) 
library(patchwork)

annot = read.csv("/users/genomics/marta/TestisProject_SaraRazquin/with_TranscriptomeReconstruction/human/newReference_Resconstructed/1transcript_1gene.reconstructed.csv")
# names(gene_transcript) = c("gene_id"gene_type"gene_name")
# gene_transcript$gene_id = gsub("\\..*"gene_transcript$gene_id)

plots_wd = "/projects_eg/projects/marta/TestisRestricted_Microproteins_TSA/with_TranscriptomeReconstruction/Q1_EvolutionaryOrigin/human/plots"
save_wd = "/projects_eg/projects/marta/TestisRestricted_Microproteins_TSA/with_TranscriptomeReconstruction/Q1_EvolutionaryOrigin/human"

species=c("mouse","macaca","platypus","chicken","opossum")

evoDir = "/users/genomics/marta/TestisProject_SaraRazquin/with_TranscriptomeReconstruction/EvolutionaryOrigin_MMseqs"

mycolors=list("gene_type" = c("protein_coding"="#CC79A7",
                               "lncRNA" = "#009E73",
                               "processed_pseudogene" = "#0090B2",
                               "novel" = "#E69F00"))
```

# Evolutionary Origin

For human, only those that are translated in Testis:

`/projects_eg/projects/marta/TestisRestricted_Microproteins_TSA/with_TranscriptomeReconstruction/Q1_TestisORFs/human/ribORF_humanTestis_in1.fa`
`/projects_eg/projects/marta/TestisRestricted_Microproteins_TSA/with_TranscriptomeReconstruction/Q1_TestisORFs/human/ribORF_humanTestis_in1.csv`

Exclusively translated in testis:

`/projects_eg/projects/marta/TestisRestricted_Microproteins_TSA/with_TranscriptomeReconstruction/Q2_TestisRestricted/human/testisRestricted_GTEx_translatedONLYtestis.csv`

Tumor Reactivated (TOv3x5%)

`/projects_eg/projects/marta/TestisRestricted_Microproteins_TSA/with_TranscriptomeReconstruction/Q4_TestisRestricted_TumorSpecific/human/TOv3x_5percent_TestisRestrictedGTEx_Translated_Ctypes.csv`

```{r read, echo=F}
testis = read.csv("/projects_eg/projects/marta/TestisRestricted_Microproteins_TSA/with_TranscriptomeReconstruction/Q1_TestisORFs/human/ribORF_humanTestis_in1.csv")
testisRestr = read.csv('/projects_eg/projects/marta/TestisRestricted_Microproteins_TSA/with_TranscriptomeReconstruction/Q2_TestisRestricted/human/testisRestricted_GTEx_translatedONLYtestis.csv')
tumorReact = read.csv("/projects_eg/projects/marta/TestisRestricted_Microproteins_TSA/with_TranscriptomeReconstruction/Q4_TestisRestricted_TumorSpecific/human/TOv3x_5percent_TestisRestrictedGTEx_Translated_Ctypes.csv")
```

For the species, everything that is translated

`/users/genomics/marta/TestisProject_SaraRazquin/with_TranscriptomeReconstruction/*/RiboSeq/RiboQC_RiboNovel*/RibORF/*_repre.valid.pred.pvalue.parameters.PROTEIN.noredunant.fa`

## Testis Translated

### Non-coding Conserved

```{r nc_conserved, echo=F}
nc_conserved = read.csv(file.path(evoDir,"Clustering/noncodingHuman_Conserved_matrix.tsv"), sep="\t")
nc_conserved_ordered = nc_conserved[order(nc_conserved$sum, decreasing = TRUE), ] 

matrix = nc_conserved_ordered %>% select(macaca, mouse, opossum, platypus, chicken)
matrix = apply(matrix, 2, as.numeric)
matrix = as.matrix(matrix)
rownames(matrix) = nc_conserved$orfID

## row annotation - gene_type
row_annotation = nc_conserved_ordered %>% select(orfID, gene_type)
rownames(row_annotation) = row_annotation$orfID
row_annotation$orfID = NULL

pheatmap(matrix, cluster_rows=FALSE, cluster_cols=FALSE,
         annotation_names_col = FALSE, 
         annotation_names_row = FALSE,
         show_colnames = TRUE,
         show_rownames= FALSE,
         annotation_row = row_annotation,
         annotation_colors = mycolors,  # Custom colors
         color = colorRampPalette(c("lightyellow", "#F57F17"))(50),
         fontsize = 6,
         main="Conservation of ncORFs in human | Testis")


#### conserved and young
all_nc = read.csv(file.path(evoDir,"Clustering/noncodingHuman_matrix.tsv"), sep="\t")
all_nc_ordered = all_nc[order(all_nc$sum, decreasing = TRUE), ] 

matrix = all_nc_ordered %>% select(macaca, mouse, opossum, platypus, chicken)
matrix = apply(matrix, 2, as.numeric)
matrix = as.matrix(matrix)
rownames(matrix) = all_nc$orfID

## row annotation - gene_type
row_annotation = all_nc_ordered %>% select(orfID, gene_type)
rownames(row_annotation) = row_annotation$orfID
row_annotation$orfID = NULL

pheatmap(matrix, cluster_rows=FALSE, cluster_cols=FALSE,
         annotation_names_col = FALSE, 
         annotation_names_row = FALSE,
         show_colnames = TRUE,
         show_rownames= FALSE,
         annotation_row = row_annotation,
         annotation_colors = mycolors,  # Custom colors
         color = colorRampPalette(c("lightyellow", "#F57F17"))(50),
         fontsize = 6,
         main="Evolutionary Origin of ncORFs in human | Testis")
```

### Canonical Conserved

```{r cds_conserved, echo=F}
cds_conserved = read.csv(file.path(evoDir,"Clustering/canonicalHuman_Conserved_matrix.tsv"), sep="\t")
cds_conserved_ordered = cds_conserved[order(cds_conserved$sum, decreasing = TRUE), ] 

matrix = cds_conserved_ordered %>% select(macaca, mouse, opossum, platypus, chicken)
matrix = apply(matrix, 2, as.numeric)
matrix = as.matrix(matrix)
rownames(matrix) = cds_conserved$orfID

## row annotation - gene_type
row_annotation = cds_conserved_ordered %>% select(orfID, gene_type)
rownames(row_annotation) = row_annotation$orfID
row_annotation$orfID = NULL

pheatmap(matrix, cluster_rows=FALSE, cluster_cols=FALSE,
         annotation_names_col = FALSE, 
         annotation_names_row = FALSE,
         show_colnames = TRUE,
         show_rownames= FALSE,
         annotation_row = row_annotation,
         annotation_colors = mycolors,  # Custom colors
         color = colorRampPalette(c("lightyellow", "#F57F17"))(50),
         fontsize = 6,
         main="Conservation of CDS in human | Testis")

#### conserved and young
all_cds = read.csv(file.path(evoDir,"Clustering/canonicalHuman_matrix.tsv"), sep="\t")
all_cds_ordered = all_cds[order(all_cds$sum, decreasing = TRUE), ] 

matrix = all_cds_ordered %>% select(macaca, mouse, opossum, platypus, chicken)
matrix = apply(matrix, 2, as.numeric)
matrix = as.matrix(matrix)
rownames(matrix) = all_cds$orfID

## row annotation - gene_type
row_annotation = all_cds_ordered %>% select(orfID, gene_type)
rownames(row_annotation) = row_annotation$orfID
row_annotation$orfID = NULL

pheatmap(matrix, cluster_rows=FALSE, cluster_cols=FALSE,
         annotation_names_col = FALSE, 
         annotation_names_row = FALSE,
         show_colnames = TRUE,
         show_rownames= FALSE,
         annotation_row = row_annotation,
         annotation_colors = mycolors,  # Custom colors
         color = colorRampPalette(c("lightyellow", "#F57F17"))(50),
         fontsize = 6,
         main="Evolutionary Origin of CDS in human | Testis")
```

## Both canonical and ncORFs conserved

```{r merged, echo=F}
conserved = rbind(cds_conserved, nc_conserved)
conserved_ordered = conserved[order(conserved$sum, decreasing = TRUE), ] 

matrix = conserved_ordered %>% select(macaca, mouse, opossum, platypus, chicken)
matrix = apply(matrix, 2, as.numeric)
matrix = as.matrix(matrix)
rownames(matrix) = conserved$orfID

## row annotation - gene_type
row_annotation = conserved_ordered %>% select(orfID, gene_type)
rownames(row_annotation) = row_annotation$orfID
row_annotation$orfID = NULL

pheatmap(matrix, cluster_rows=FALSE, cluster_cols=FALSE,
         annotation_names_col = FALSE, 
         annotation_names_row = FALSE,
         show_colnames = TRUE,
         show_rownames= FALSE,
         annotation_row = row_annotation,
         annotation_colors = mycolors,  # Custom colors
         color = colorRampPalette(c("lightyellow", "#F57F17"))(50),
         fontsize = 6,
         main="Conservation of ORFs in human")

merged = rbind(all_nc, all_cds)

#### conserved and young
merged_ordered = merged[order(merged$sum, decreasing = TRUE), ] 

matrix = merged_ordered %>% select(macaca, mouse, opossum, platypus, chicken)
matrix = apply(matrix, 2, as.numeric)
matrix = as.matrix(matrix)
rownames(matrix) = merged_ordered$orfID

## row annotation - gene_type
row_annotation = merged_ordered %>% select(orfID, gene_type)
rownames(row_annotation) = row_annotation$orfID
row_annotation$orfID = NULL

pheatmap(matrix, cluster_rows=F, cluster_cols=FALSE,
         annotation_names_col = FALSE, 
         annotation_names_row = FALSE,
         show_colnames = TRUE,
         show_rownames= FALSE,
         annotation_row = row_annotation,
         annotation_colors = mycolors,  # Custom colors
         color = colorRampPalette(c("lightyellow", "#F57F17"))(50),
         fontsize = 6,
         main="Evolutionary Origin of CDS and ncORFs in human | Testis")
```

```{r age_merged_all, echo=F}
evoAge = read.csv("/users/genomics/marta/TestisProject_SaraRazquin/with_TranscriptomeReconstruction/EvolutionaryOrigin_MMseqs/Clustering/allHuman_matrix_oldestAge.tsv", sep="\t")

testis_age = evoAge %>% subset(orfID %in% merged_ordered$orfID)
testis_age_count = testis_age %>% group_by(gene_type, oldest_species, age) %>% count()
testis_age_count = testis_age_count %>% group_by(gene_type) %>% mutate(proportion = n / sum(n))
ggplot(testis_age_count, aes(x=as.factor(age), y=n, color=gene_type)) +
  geom_point() +
  geom_line(aes(group=gene_type)) +
  scale_color_manual(values = c("protein_coding"="#CC79A7",
                               "lncRNA" = "#009E73",
                               "processed_pseudogene" = "#0090B2")) +
  labs(x="Myr",
       y="Number of ORFs",
       title="Evolutionary age of conservation of translation\nTestis Translated") +
  theme_minimal()

ggplot(testis_age_count, aes(x=as.factor(age), y=proportion, color=gene_type)) +
  geom_point() +
  geom_line(aes(group=gene_type)) +
  scale_color_manual(values = c("protein_coding"="#CC79A7",
                               "lncRNA" = "#009E73",
                               "processed_pseudogene" = "#0090B2")) +
  labs(x="Myr",
       y="Number of ORFs",
       title="Evolutionary age of conservation of translation\nTestis Translated") +
  theme_minimal()
```
## Testis-Restricted Translated

```{r testisRestricted, echo=F}
TRestr_ordered = merged_ordered %>% subset(orfID %in% testisRestr$orfID)
TRestr_ordered_conserved = conserved_ordered %>% subset(orfID %in% testisRestr$orfID)

matrix = TRestr_ordered %>% select(macaca, mouse, opossum, platypus, chicken)
matrix = apply(matrix, 2, as.numeric)
matrix = as.matrix(matrix)
rownames(matrix) = TRestr_ordered$orfID

## row annotation - gene_type
row_annotation = TRestr_ordered %>% select(orfID, gene_type)
rownames(row_annotation) = row_annotation$orfID
row_annotation$orfID = NULL

pheatmap(matrix, cluster_rows=FALSE, cluster_cols=FALSE,
         annotation_names_col = FALSE, 
         annotation_names_row = FALSE,
         show_colnames = TRUE,
         show_rownames= FALSE,
         annotation_row = row_annotation,
         annotation_colors = mycolors,  # Custom colors
         color = colorRampPalette(c("lightyellow", "#F57F17"))(50),
         fontsize = 6,
         main="Conservation of ncORFs & CDSs in human | TestisRestricted")

### only conserved
matrix = TRestr_ordered_conserved %>% select(macaca, mouse, opossum, platypus, chicken)
matrix = apply(matrix, 2, as.numeric)
matrix = as.matrix(matrix)
rownames(matrix) = TRestr_ordered_conserved$orfID

## row annotation - gene_type
row_annotation = TRestr_ordered_conserved %>% select(orfID, gene_type)
rownames(row_annotation) = row_annotation$orfID
row_annotation$orfID = NULL

pheatmap(matrix, cluster_rows=FALSE, cluster_cols=FALSE,
         annotation_names_col = FALSE, 
         annotation_names_row = FALSE,
         show_colnames = TRUE,
         show_rownames= FALSE,
         annotation_row = row_annotation,
         annotation_colors = mycolors,  # Custom colors
         color = colorRampPalette(c("lightyellow", "#F57F17"))(50),
         fontsize = 6,
         main="Conservation of ncORFs in human | TestisRestricted")
```

```{r age_testisRestr, echo=F}
testisRestr_age = evoAge %>% subset(orfID %in% TRestr_ordered$orfID)
testisRestr_age_count = testisRestr_age %>% group_by(gene_type, oldest_species, age) %>% count()
testisRestr_age_count = testisRestr_age_count %>% group_by(gene_type) %>% mutate(proportion = n / sum(n))
ggplot(testisRestr_age_count, aes(x=as.factor(age), y=n, color=gene_type)) +
  geom_point() +
  geom_line(aes(group=gene_type)) +
  scale_color_manual(values = c("protein_coding"="#CC79A7",
                               "lncRNA" = "#009E73",
                               "processed_pseudogene" = "#0090B2")) +
  labs(x="Myr",
       y="Number of ORFs",
       title="Evolutionary age of conservation of translation\nTestis-Restricted") +
  theme_minimal()

ggplot(testisRestr_age_count, aes(x=as.factor(age), y=proportion, color=gene_type)) +
  geom_point() +
  geom_line(aes(group=gene_type)) +
  scale_color_manual(values = c("protein_coding"="#CC79A7",
                               "lncRNA" = "#009E73",
                               "processed_pseudogene" = "#0090B2")) +
  labs(x="Myr",
       y="Number of ORFs",
       title="Evolutionary age of conservation of translation\nTestis-Restricted") +
  theme_minimal()
```

## Testis-Restricted Translated Reactivated in Tumor (3x Normal in 5%)

```{r tumorReactivated, echo=F}
tumorReactivated_ordered = merged_ordered %>% subset(orfID %in% tumorReact$orfID)
tumorReactivated_ordered_conserved = conserved_ordered %>% subset(orfID %in% tumorReact$orfID)

matrix = tumorReactivated_ordered %>% select(macaca, mouse, opossum, platypus, chicken)
matrix = apply(matrix, 2, as.numeric)
matrix = as.matrix(matrix)
rownames(matrix) = tumorReactivated_ordered$orfID

## row annotation - gene_type
row_annotation = tumorReactivated_ordered %>% select(orfID, gene_type)
rownames(row_annotation) = row_annotation$orfID
row_annotation$orfID = NULL

pheatmap(matrix, cluster_rows=FALSE, cluster_cols=FALSE,
         annotation_names_col = FALSE, 
         annotation_names_row = FALSE,
         show_colnames = TRUE,
         show_rownames= FALSE,
         annotation_row = row_annotation,
         annotation_colors = mycolors,  # Custom colors
         color = colorRampPalette(c("lightyellow", "#F57F17"))(50),
         fontsize = 6,
         main="Conservation of ncORFs & CDSs in human | TestisRestricted Reactivated in Tumors")

### only conserved
matrix = tumorReactivated_ordered_conserved %>% select(macaca, mouse, opossum, platypus, chicken)
matrix = apply(matrix, 2, as.numeric)
matrix = as.matrix(matrix)
rownames(matrix) = tumorReactivated_ordered_conserved$orfID

## row annotation - gene_type
row_annotation = tumorReactivated_ordered_conserved %>% select(orfID, gene_type)
rownames(row_annotation) = row_annotation$orfID
row_annotation$orfID = NULL

pheatmap(matrix, cluster_rows=FALSE, cluster_cols=FALSE,
         annotation_names_col = FALSE, 
         annotation_names_row = FALSE,
         show_colnames = TRUE,
         show_rownames= FALSE,
         annotation_row = row_annotation,
         annotation_colors = mycolors,  # Custom colors
         color = colorRampPalette(c("lightyellow", "#F57F17"))(50),
         fontsize = 6,
         main="Conservation of ncORFs in human | TestisRestricted Reactivated in Tumors")
```

```{r age_tumorReact, echo=F}
tumorReact_age = evoAge %>% subset(orfID %in% tumorReactivated_ordered$orfID)
tumorReact_age_count = tumorReact_age %>% group_by(gene_type, oldest_species, age) %>% count()
tumorReact_age_count = tumorReact_age_count %>% group_by(gene_type) %>% mutate(proportion = n / sum(n))
ggplot(tumorReact_age_count, aes(x=as.factor(age), y=n, color=gene_type)) +
  geom_point() +
  geom_line(aes(group=gene_type)) +
  scale_color_manual(values = c("protein_coding"="#CC79A7",
                               "lncRNA" = "#009E73",
                               "processed_pseudogene" = "#0090B2")) +
  labs(x="Myr",
       y="Number of ORFs",
       title="Evolutionary age of conservation of translation\nTestis-Restricted Reactivated in Tumors") +
  theme_minimal()

ggplot(tumorReact_age_count, aes(x=as.factor(age), y=proportion, color=gene_type)) +
  geom_point() +
  geom_line(aes(group=gene_type)) +
  scale_color_manual(values = c("protein_coding"="#CC79A7",
                               "lncRNA" = "#009E73",
                               "processed_pseudogene" = "#0090B2")) +
  labs(x="Myr",
       y="Number of ORFs",
       title="Evolutionary age of conservation of translation\nTestis-Restricted Reactivated in Tumors") +
  theme_minimal()
```

### Ages summary

```{r ages, echo=F}
tumorReact_age$condition = "TumorReactivated"
testisRestr_age$condition = "TestisRestricted"
testis_age$condition = "Testis"


ages = rbind(tumorReact_age, testisRestr_age, testis_age)
ages = ages %>% mutate(age_categorical = case_when(age < 100 ~ "young",
                                                   TRUE ~ "old"))

ages_count = ages %>% group_by(gene_type, age_categorical, condition) %>% count()
ages_count = ages_count %>% group_by(gene_type, condition) %>% mutate(proportion = n / sum(n))
# ggplot(ages_count, aes(x=as.factor(age), y=n, color=gene_type)) +
#   geom_point() +
#   geom_line(aes(group=gene_type)) +
#   scale_color_manual(values = c("protein_coding"="#CC79A7",
#                                "lncRNA" = "#009E73",
#                                "processed_pseudogene" = "#0090B2")) +
#   labs(x="Myr",
#        y="Number of ORFs",
#        title="Evolutionary age of conservation of translation\nTestis-Restricted Reactivated in Tumors") +
#   theme_minimal() +
#   theme(legend.position="top")+
#   facet_wrap(~ condition, scales="free_y")
# 
# ggplot(ages_count, aes(x=as.factor(age), y=proportion, color=gene_type)) +
#   geom_point() +
#   geom_line(aes(group=gene_type)) +
#   scale_color_manual(values = c("protein_coding"="#CC79A7",
#                                "lncRNA" = "#009E73",
#                                "processed_pseudogene" = "#0090B2")) +
#   labs(x="Myr",
#        y="Number of ORFs",
#        title="Evolutionary age of conservation of translation\nTestis-Restricted Reactivated in Tumors") +
#   theme_minimal() +
#   theme(legend.position="top")+
#   facet_wrap(~ condition, scales="free_y")

ggplot(ages_count, aes(x=factor(age_categorical, levels = c("young","old")), y=n, color=gene_type)) +
  geom_point() +
  geom_line(aes(group=gene_type)) +
  scale_color_manual(values = c("protein_coding"="#CC79A7",
                               "lncRNA" = "#009E73",
                               "processed_pseudogene" = "#0090B2")) +
  labs(x="Myr",
       y="Number of ORFs",
       title="Evolutionary age of conservation of translation\nTestis-Restricted Reactivated in Tumors") +
  theme_minimal() +
  theme(legend.position="top")+
  facet_wrap(~ condition, scales="free_y")

ggplot(ages_count, aes(x=factor(age_categorical, levels = c("young","old")), y=proportion, color=gene_type)) +
  geom_point() +
  geom_line(aes(group=gene_type)) +
  scale_color_manual(values = c("protein_coding"="#CC79A7",
                               "lncRNA" = "#009E73",
                               "processed_pseudogene" = "#0090B2")) +
  labs(x="Myr",
       y="Number of ORFs",
       title="Evolutionary age of conservation of translation\nTestis-Restricted Reactivated in Tumors") +
  theme_minimal() +
  theme(legend.position="top")+
  facet_wrap(~ condition, scales="free_y")
```

## Length

```{r lengths, echo=F}
testis_length = testis %>% select(orfID, gene_type, length_aa) %>% mutate(condition = "Testis")
testisRestr_length = testisRestr %>% select(orfID, gene_type, length_aa) %>% mutate(condition = "TestisRestricted")
tumorReact_length = tumorReact %>% select(orfID, gene_type, length_aa) %>% mutate(condition = "TumorReactivated")

lengths = rbind(testis_length, testisRestr_length, tumorReact_length)

ages_lengths = merge(lengths, ages, by=c("orfID","gene_type","condition"))
ages_lengths = ages_lengths %>% mutate(age_categorical = case_when(age < 100 ~ "young",
                                                                   TRUE ~ "old"))

ggplot(ages_lengths, aes(x=length_aa, color=gene_type)) +
  geom_density() +
  scale_x_continuous(trans="log10") +
  scale_color_manual(values = c("protein_coding"="#CC79A7",
                             "lncRNA" = "#009E73",
                             "processed_pseudogene" = "#0090B2")) +
  labs(x="Length (AA)",
       title="Length distribution") +
  theme_minimal() +
  theme(legend.position = "top") +
  facet_wrap(~ condition, scales="free_y")

ggplot(ages_lengths, aes(x=as.factor(age), y=length_aa, fill=gene_type)) +
  geom_boxplot() +
  scale_fill_manual(values = c("protein_coding"="#CC79A7",
                             "lncRNA" = "#009E73",
                             "processed_pseudogene" = "#0090B2")) +
  scale_y_continuous(trans="log10") +
    theme_minimal() +
  theme(legend.position = "top") +
  facet_wrap(~ condition, scales="free_y")

ggplot(ages_lengths, aes(x=factor(age_categorical, levels=c("young","old")), y=length_aa, fill=gene_type)) +
  geom_boxplot() +
  scale_fill_manual(values = c("protein_coding"="#CC79A7",
                             "lncRNA" = "#009E73",
                             "processed_pseudogene" = "#0090B2")) +
  scale_y_continuous(trans="log10") +
    theme_minimal() +
  theme(legend.position = "top") +
  facet_wrap(~ condition, scales="free_y")

ages_lengths_median = ages_lengths %>% group_by(gene_type, age, age_categorical, condition) %>% mutate(median_length = median(length_aa))
ggplot(ages_lengths_median, aes(x=as.factor(age), y=median_length, color=gene_type)) +
  geom_point() +
  geom_line(aes(group=gene_type)) +
  scale_color_manual(values = c("protein_coding"="#CC79A7",
                             "lncRNA" = "#009E73",
                             "processed_pseudogene" = "#0090B2")) +
  scale_y_continuous(trans="log10") +
    theme_minimal() +
  theme(legend.position = "top") +
  facet_wrap(~ condition, scales="free_y")

ggplot(ages_lengths_median, aes(x=as.factor(age), y=median_length, color=gene_type)) +
  geom_point() +
  geom_line(aes(group=gene_type)) +
  scale_color_manual(values = c("protein_coding"="#CC79A7",
                             "lncRNA" = "#009E73",
                             "processed_pseudogene" = "#0090B2")) +
  scale_y_continuous(trans="log10") +
    theme_minimal() +
  theme(legend.position = "top") +
  facet_wrap(~ condition, scales="free_y")
```

## TE

```{r TE, echo=F}
mean_testis_TE = read.csv("/projects_eg/projects/marta/TestisRestricted_Microproteins_TSA/with_TranscriptomeReconstruction/Q1_TestisORFs/human/mean_translationEfficiency_testis.csv")
testis_TE = mean_testis_TE %>% select(orfID, gene_type, mean_TE) %>% subset(orfID %in% testis$orfID) %>% mutate(condition = "Testis")
testisRestr_TE = mean_testis_TE %>% select(orfID, gene_type, mean_TE) %>% subset(orfID %in% testisRestr$orfID) %>% mutate(condition = "TestisRestricted")
tumorReact_TE = mean_testis_TE %>% select(orfID, gene_type, mean_TE) %>% subset(orfID %in% tumorReact$orfID) %>% mutate(condition = "TumorReactivated")

TE = rbind(testis_TE, testisRestr_TE, tumorReact_TE)
TE_ages_length = merge(TE, ages_lengths, by=c("gene_type","condition","orfID"))

ggplot(TE, aes(x=condition, y=mean_TE, fill=gene_type)) +
  geom_boxplot() +
  scale_fill_manual(values = c("protein_coding"="#CC79A7",
                             "lncRNA" = "#009E73",
                             "processed_pseudogene" = "#0090B2")) +
  scale_y_continuous(trans="log10") +
  labs(x="",
       y="Translation Efficiency in testis (mean)",
       title="Translation Efficiency") +
  theme_minimal() +
  theme(legend.position = "top") 

```

### yes - no 

```{r proportions_yes_no, echo=F}
all_info = TE_ages_length %>% subset(condition == "Testis") %>% select(-condition)
all_info = all_info %>% mutate(testisRestricted = case_when(orfID %in% testisRestr$orfID ~ "yes",
                                                            TRUE ~ "no"),
                               tumorReactivated = case_when(orfID %in% tumorReact$orfID ~ "yes",
                                                            TRUE ~ "no"))
```

#### Length 

```{r yes_no_length, echo=F}
ggplot(all_info, aes(x=testisRestricted, y=length_aa, color=gene_type)) +
  geom_boxplot() +
  scale_y_continuous(trans="log10") +
    stat_compare_means(aes(group=testisRestricted), label="p.signif") +
  scale_color_manual(values = c("protein_coding"="#CC79A7",
                             "lncRNA" = "#009E73",
                             "processed_pseudogene" = "#0090B2")) +
  labs(y="Length (AA)",
       title="Length differences for testisRestricted ORFs") +
  theme_minimal() +
  theme(legend.position = "top") +
  facet_wrap(~ gene_type) 

ggplot(all_info, aes(x=tumorReactivated, y=length_aa, color=gene_type)) +
  geom_boxplot() +
  scale_y_continuous(trans="log10") +
    stat_compare_means(aes(group=tumorReactivated), label="p.signif") +
  scale_color_manual(values = c("protein_coding"="#CC79A7",
                             "lncRNA" = "#009E73",
                             "processed_pseudogene" = "#0090B2")) +
  labs(y="Length (AA)",
       title="Length differences for tumorReactivated ORFs") +
  theme_minimal() +
  theme(legend.position = "top") +
  facet_wrap(~ gene_type) 
```

#### TE 

```{r yes_no_TE, echo=F}
ggplot(all_info, aes(x=testisRestricted, y=mean_TE, color=gene_type)) +
  geom_boxplot() +
  scale_y_continuous(trans="log10") +
    stat_compare_means(aes(group=testisRestricted), label="p.signif") +
  scale_color_manual(values = c("protein_coding"="#CC79A7",
                             "lncRNA" = "#009E73",
                             "processed_pseudogene" = "#0090B2")) +
  labs(y="Translation Efficiency in Testis (mean)",
       title="Translation Efficiency for testisRestricted ORFs") +
  theme_minimal() +
  theme(legend.position = "top") +
  facet_wrap(~ gene_type) 

ggplot(all_info, aes(x=tumorReactivated, y=mean_TE, color=gene_type)) +
  geom_boxplot() +
  scale_y_continuous(trans="log10") +
    stat_compare_means(aes(group=tumorReactivated), label="p.signif") +
  scale_color_manual(values = c("protein_coding"="#CC79A7",
                             "lncRNA" = "#009E73",
                             "processed_pseudogene" = "#0090B2")) +
  labs(y="Translation Efficiency in Testis (mean)",
       title="Translation Efficiency for tumorReactivated ORFs") +
  theme_minimal() +
  theme(legend.position = "top") +
  facet_wrap(~ gene_type) 
```

#### "Evolutionary Age"

```{r yes_no_age, echo=F}
```